### 2021-03-01 paper.Rmd
add alternatives

A tour can be considered to be a dynamic graphic, because it shows a smooth sequence of projections over time, ideally with controls that allow stopping, reversing, changing direction, or going forward again. It can be useful to embed a tour into an interactive graphics system, where plots can be queried and elements highlighted (see for e.g. @Swayne2003-cc or @Tierney1991). To create the smooth sequence, a geodesic interpolation is computed between consecutive frames.  It allows the viewer to extrapolate from the low-dimensional to  shapes corresponding to multivariate distribution, and is particularly useful for detecting clusters, outliers and non-linear dependence.

<!-- Original tours (Di)
- Dataviewer, XGobi, GGobi, cranvas (XLispStat, Orca, mmvz) (Di)
(picture of the cluster data from Stuart's thesis)
-->
- Alternatives (Stuart)

(parallel coordinates, sploms, what tours are not)

@@

- Model vis examples eg random forest (from Hadley, removing the blindfold) (Natalia)
- Physics (Ursula)
- Bioinformatics (Stuart)
- High-d shapes/distributions: normal, skewed, outliers, (Di from tutorial notes)

>>>>

A tour can be considered to be a dynamic graphic, because it shows a smooth sequence of projections over time, ideally with controls that allow stopping, reversing, changing direction, or going forward again. It can be useful to embed a tour into an interactive graphics system, where plots can be queried and elements highlighted (see for e.g. @Swayne2003-cc or @Tierney1991). To create the smooth sequence, a geodesic interpolation is computed between consecutive frames.  It allows the viewer to extrapolate from the low-dimensional to  shapes corresponding to multivariate distribution, and is particularly useful for detecting clusters, outliers and non-linear dependence. 
While tours are invaluable for assessing the geometry of data, they are by no means the only technique available for visualising structure in high dimensional data. An early technique proposed for assessing pairwise relationships between variables is the scatter plot matrix (SPLOM) [@Tukey1983-fj] <!-- there might be others here(?) -->. In this graphic, the canvas is split into a matrix where each off-diagonal element is a scatter plot of variable pairs, while the diagonal elements are either left blank or represent univariate structure. The SPLOM allows the viewer to assess correlation structure but does not scale to large numbers of variables.

Similarly, parallel coordinates plots (PCP) can be used to explore correlation and co-linearities in multivariate data [@Inselberg1985-kf]. In PCPs, variables are arranged along the (vertical) horizontal axis with vertical (horizontal) axes drawn representing the range of values for each variable. Points (observations) of the data are represented by drawing lines between each variables axis. If there is positive correlation between two variables the lines will be parallel, while for negative correlation the lines will cross. By placing multiple variables side by side in a PCP higher order structure like clustering or lower dimensional embeddings is revealed, however the ordering of variables is important to reveal these features. <!-- anything else? Andrew's curves etc.? -->

<!-- Original tours (Di)
- Dataviewer, XGobi, GGobi, cranvas (XLispStat, Orca, mmvz) (Di)
(picture of the cluster data from Stuart's thesis)
- Alternatives (Stuart) 
-->

@@

- Model vis examples eg random forest (from Hadley, removing the blindfold) (Natalia)
- Physics (Ursula)
- Bioinformatics (Stuart)

- High-d shapes/distributions: normal, skewed, outliers, (Di from tutorial notes)
